---
title: "Class09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hands On

```{r}
# Save your input data file to a new 'data' directory
fna.data <-  "WisconsinCancer.csv"
```

```{r}
# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data)
```
Lets see the first six rows of data 
```{r}
head(wisc.df)
```
```{r}
View(wisc.df)
```
How many samples (i.e. patients) are in this data set? 
```{r}
nrow(wisc.df)
```
Next use as.matrix() to convert the other features (i.e. columns) of the data (in columns 3 through 32) to a matrix. Store this in a variable called wisc.data.
```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc.df[,3:32])
```

Add rownames to our new matric of data

```{r}
row.names(wisc.data) <- wisc.df$id
```
Finally, setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and 0 otherwise. Note that R coerces TRUE to 1 and FALSE to 0.
```{r}
wisc.df$diagnosis
```

How many cancer (M) and non cancer samples do we have in our data set?
```{r}
 table(wisc.df$diagnosis)
```
```{r}
#diagnosis <- as.numeric(wisc.df$diagnosis)
#tmp <- rep(0, nrow(wisc.df))
diagnosis<- as.numeric(wisc.df$diagnosis == "M")
```
Hands on Questions
Q1. How many observations are in this dataset?
```{r}
nrow(wisc.data)
```

Q2. How many variables/features in the data are
suffixed with _mean?
```{r}
 grep("_mean", colnames(wisc.data))

```

Q3. How many of the observations have a malignant diagnosis?
```{r}
 sum (diagnosis)
```
#Section 2 

Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the colMeans() and apply() functions like you’ve done before.
```{r}
# Check column means and standard deviations
 colMeans(wisc.data)

 apply(wisc.data,2,sd)
```

Now lets do PCA
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale= TRUE)
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
```{r}
looking at the table above 0.4427

```

Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
```{r}
you would look at the proportion of variance above and realize you need to sum PC1-PC5 so 5 components
```

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
```{r}
seven 
```
We need to make our own PCA plot 
btw 1 is white and 2 is red so +1 gives red
```{r}
plot( wisc.pr$x[, 1], wisc.pr$x[,2])
plot(wisc.pr$x[, 1], wisc.pr$x[,2], col=diagnosis+1)
plot(wisc.pr$x[, 1], wisc.pr$x[,2], col=diagnosis+1,
     xlab= "PC1", ylab="PC2")
```
What if we wanted to plot PC1 vs PC3?
```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[,3], col=diagnosis+1,
     xlab= "PC1", ylab="PC3")
```



This info is in the $sdev component of our PCA result
```{r}
varance <- wisc.pr$sdev^2
# Variance explained by each principal component: pve
pve <- round ( (varance/ sum(varance)) *100,1 )
```
```{r}
plot(pve, type= "o")
```


Make the plot a bit fancy :-o

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 ) 


#axes=false turns of the axes so that you can reset them after in the following line of code
```

Section 3 Hands On
## Hierarchial clustering of cancer case data
For hierarchial clustering we need a few things
1. Distance matrix **dist()** function
2. The **hclust()** function 
3. Use the **cutree()** function to find cluster membership vector

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```
```{r}
round( apply(wisc.data,2, sd),1)
```
```{r}
round (apply(data.scaled, 2, sd), 1)
```

Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.
```{r}
data.dist <- dist(data.scaled)
```
Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign the results to wisc.hclust.
```{r}
wisc.hclust <- hclust(data.dist, method= "complete" )
wisc.hclust
```


```{r}
plot(wisc.hclust)
```

Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
Hide
```{r}
 wisc.pr.hclust <- plot(wisc.hclust)
h <- "height"
abline(h=19, col="red", lty=2)
```
Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters.

Section 5
## Cluster in PCA space

we need
1. distance matrix
2. clustering function
3. cutree

This was our PCA resut of PC1 vs Pc2
```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[,2], col=diagnosis+1,
     xlab= "PC1", ylab="PC2")
```
```{r}
pc.dist <- dist(wisc.pr$x[,1:2])
pc.hclust <- hclust(pc.dist, method="ward.D2")
plot(pc.hclust)
```

```{r}
grps3 <- cutree(pc.hclust, k=3, h= 19)
grps3
```
```{r}
#how many in each cluster
table(grps3)
```
```{r}
table(grps3, diagnosis)
```
Section 7
```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
The 2 blue points are 2 new patients, new data recorded as part of the new data set above
```{r}
plot(wisc.pr$x[,1:2], col=diagnosis + 1)
points(npc[,1], npc[,2], col="blue", pch=15, cex=3)

```













